[Re] Predicting metabolic biomarkers of human inborn errors of metabolism

### 2018-02-15 article/Mondeel_Ogundipe_Westerhoff-2017.tex
small changes
- Vivian has moved to Groningen
- flux bounds of -1 not 1
- fix missing table reference
- consistently report we reproduce fig. 1 through 3.

    \usepackage{authblk}
                        \author[1,*]{Thierry D.G.A. Mondeel}
                    \author[1]{Vivian Ogundipe}
                    \author[1, 2, 3]{Hans V. Westerhoff}
                            \affil[1]{Synthetic Systems Biology and Nuclear Organization, Swammerdam Institute
for Life Sciences, University of Amsterdam, The Netherlands}
                    \affil[2]{Molecular Cell Physiology, Amsterdam Institute for Molecules, Medicines
and Systems, VU University Amsterdam, The Netherlands}
                    \affil[3]{Manchester Centre for Integrative Systems Biology, School of Chemical
Engineering and Analytical Science, The University of Manchester, UK}

@@

Each biomarker will be associated with a prediction for being either
elevated or decreased in the mutant case, as comparted to the wild type.
A biomarker is considered to have an increased, or reduced,
extracellular concentration in the mutant case if, when plotting the
wild-type and mutant flux-variability intervals on a horizontal axis,
both the lower and upper boundary of the mutant interval are shifted to
the right, or left, respectively (see Figures
~\ref{fig:Shlomi2009_figure_1}B and ~\ref{fig:Shlomi2009_figure_3}B). If
the two borders of the mutant interval move in opposite directions with
respect to the wild-type interval, the result is scored as `unchanged'
and the boundary metabolite is not considered to be a biomarker.

@@

Here \(\bm{v}\) is the column vector of fluxes representing all
reactions in the model, \(v_i\) is the flux of the exchange reaction of
a given boundary metabolite \(i\), \(\alpha_k\) is the, possibly
negative, lower bound for reaction \(k\) and similarly \(\beta_k\) is
the, possibly negative, upper bound for reaction \(k\). The reaction
bounds are the \(V_{max}\)'s of the reactions and are part of the
metabolic network definition, typically they are set to either \(0\) or
\(\pm 1000\) if the true, biological, \(V_{max}\)'s are not known. These
values allow specification of irreversible reactions by setting the
lower bound (or upper bound) to zero. The index \(i\) specifies a single
reaction that is to be maximized, whereas the index \(k\) is used to
index the bound constraints that exist for each reaction, including
\(v_i\). \(v_i\) may be maximized in the forward (positive flux) or the
reverse (negative flux) direction if allowed by the bounds on \(v_i\).
This is indicated by the \(\pm\) symbol. \(\bm{S}\) is the stoichiometry
matrix for the metabolic network of size \(m \times r\) where \(m\) is
the number of metabolites and \(r\) the number of reactions. The
numeric, typically integer, elements of \(\bm{S}\) represent the
stoichiometry coefficients of each metabolite \(m\) in each reaction
\(r\). Consequently, for large networks, this matrix is sparse (has many
zero entries). \(Z = \bm{c}^T \bm{v}\) is the objective function defined
for the map, entailing a linear combination, defined by column vector
\(\bm{c}\), of one or more reactions. The vector \(\bm{c}\) is an
indicator of objective reactions, i.e.~it contains a value of \(1\) in
rows corresponding to fluxes that are to be included in the objective,
and zeros in all other rows. Due to the vector multiplication
\(\bm{c}^T \bm{v}\), the objective \(Z\) sums the fluxes of the
reactions that correspond to rows containing a value of \(1\) in
\(\bm{c}\). \(Z_{\text{max}}\) denotes the maximal value of the
objective function. We define \(0 \leq \phi \leq 1\) to be an arbitrary
minimal fraction of the maximal objective function value that has to be
achieved. When \(\phi=1\) we ask for the range of flux allowed for
reaction \(v_i\) while maintaining the maximal value of the linear
combination of objective fluxes. When setting \(\phi = 0\), one is
asking for the allowable flux range through reaction \(v_i\) regardless
of the value of the objective function (although the sign has to be
maintained). The latter, i.e. \(\phi=0\), is the case considered in the
method by Shlomi et al. The choice of \(\phi\) matters because when
\(phi \neq 0\) we are requiring flux to flow through the set of
objective reactions. This may entail a forced directionality through
reactions that are required to ultimately produce flux in the objective
reactions. These additional limitation in freedom of the flux pattern
may subsequently impact the biomarker predictions. The approach
discussed here entails predictions purely based on the network topology
and therefore sets \(\phi = 0\).

@@

We here present an implementation of the biomarker prediction method
originally proposed by Shlomi et al., programmed in Python, and
reproduce the results presented in Figure 1A-B and 2 of the original
publication. These figures concern the method's application to a simple
metabolic map and then to a set of \(17\) amino acid metabolism
disorders. Our Python module provides the biomarker prediction algorithm
and is agnostic of the map used. It can therefore be applied to other
maps than the human metabolic reconstruction considered by Shlomi et al.
Jupyter notebooks are provided that generate the illustrative map and
reproduce all results referred to in Figures 1A-B and 2 of Shlomi et al.
Our implementation is based on information given in the original paper.
However, some details critical to the reproduction were left unexplained
there and are highlighted here. We also include a more detailed
discussion of one of the amino acid IEMs to examine the robustness of
the predictions made using this approach. We feel this adds to the
utility of this reference implementation and the understanding of the
approach introduced by Shlomi et al.

@@

In this section we start with an analysis of the sensitivity of the
method which was not clearly discussed in the original publication but
is easily facilitated with our reference implementation. Subsequently,
we discuss the reproduction of Figures 1 and 2 from the original
publication.

@@

PKU patients are known for elevated levels of phenylalanine and
decreased levels of tyrosine in their serum. We illustrate this briefly
in an example also considered in Figure 2 of Shlomi et al. In Figure
~\ref{fig:PKU_network}, we sketched some essential features of the
metabolic network surrounding the phenylalanine hydroxylase (PAH) enzyme
that malfunctions in PKU. Unless stated otherwise, the simulations are
performed with a medium that allows efflux through all exchange
reactions with an upper bound of \(1000\) and influx through all
exchange reactions with a lower bound of \(1\) and we use the default
settings, including a 10\% significance threshold. The only considered
biomarkers are the \(20\) proteinogenic amino acids. We will consider
six cases in total. The results are summarized in Table
~\ref{tbl:PKU_results}.

@@

The network in Figure ~\ref{fig:Shlomi2009_figure_1}A consists of seven
metabolites, \(6\) of which (i.e.~all except M3) have exchange
reactions. We consider a hypothetical deletion of the enzyme catalyzing
the conversion of M1 into M2 coupled to the conversion of M3 into M4.
Shlomi et al. graphically provide the biomarker prediction for this
network in their Figure 1A. Here, we do so graphically and numerically
in Figure ~\ref{fig:Shlomi2009_figure_1}B and this paper is accompanied
by a Jupyter notebook reproducing the biomarker prediction. In our
analysis, all exchange reactions were given inward bounds of \(-10\) and
outward bounds of \(+1000\) so that the internal reaction bounds (of
\(\pm 1000\)) are never reached. Our Table ~\ref{tbl:toy_model_results}
shows that all exchangeable metabolites except M5 and M7 are predicted
to be robust biomarkers.

@@

It is worthwhile to note, with an eye on the sensitivity of this
approach to the various settings, that we needed to set the medium
composition very specifically. Shlomi et al. did not report on their
chosen medium in their publication but we were able to reproduce their
results by setting all inward flux bounds to \(1\) and all outward flux
bounds to \(1000\).

@@

\begin{figure}
\centering
\includegraphics{Figures/Reproduction_Shlomi_fig_3}
\caption{Reproduction of Figure 3 in \autocite{Shlomi2009}. (A)
Illustration of the subnetwork of Recon 1 relevant to the effect of
homocystinuria on the metabolism and transport of methionine. Nodes
represent metabolites and edges represent reactions. For simplicity,
only abbreviations of metabolite names are given in the Figure. The full
names are: L-Methionine (met), S-Adenosyl-L-methionine (amet),
S-Adenosyl-L-homocysteine (ahcys), L-Homocysteine (hcys), L-Serine
(ser), L-Cystathionine (cyst), L-Cysteine (cys),
S-Adenosylmethioninamine (ametam), 5-Methylthioadenosine (5mta),
5-Methylthio-5-deoxy-D-ribose 1-phosphate (5mdr1p),
5-Methylthio-5-deoxy-D-ribulose 1-phosphate (dmdru1p),
2,3-diketo-5-methylthio-1-phosphopentane (dkmpp),
2-keto-4-methylthiobutyrate (2kmb). Metabolites with a green background
color participate in additional reactions not shown here. Reactions are
indicated by their identifiers in Recon 1 and with a corresponding
enzyme commision number if available. We highlighted the mutations for
Homocystinuria (dysfunctional CBS) and hypermethioninemia (dysfunctional
AHCY). (B) Prediction of concentration changes of methionine in
homocystinuria and S-adenosylhomocysteine hydrolase deficiency based on
the biomarker prediction algorithm.}\label{fig:Shlomi2009_figure_3}
\end{figure}

>>>>

    \usepackage{authblk}
                        \author[1,*]{Thierry D.G.A. Mondeel}
                    \author[2]{Vivian Ogundipe}
                    \author[1, 3, 4]{Hans V. Westerhoff}
                            \affil[1]{Synthetic Systems Biology and Nuclear Organization, Swammerdam Institute
for Life Sciences, University of Amsterdam, The Netherlands}
                    \affil[2]{Department of Cell Biology, University Medical Center Groningen,
University of Groningen, The Netherlands}
                    \affil[3]{Molecular Cell Physiology, Amsterdam Institute for Molecules, Medicines
and Systems, VU University Amsterdam, The Netherlands}
                    \affil[4]{Manchester Centre for Integrative Systems Biology, School of Chemical
Engineering and Analytical Science, The University of Manchester, UK}

@@

Each biomarker will be associated with a prediction for being either
elevated or decreased in the mutant case, as compared to the wild type.
A biomarker is considered to have an increased, or reduced,
extracellular concentration in the mutant case if, when plotting the
wild-type and mutant flux-variability intervals on a horizontal axis,
both the lower and upper boundary of the mutant interval are shifted to
the right, or left, respectively (see Figures
~\ref{fig:Shlomi2009_figure_1}B and ~\ref{fig:Shlomi2009_figure_3}B). If
the two borders of the mutant interval move in opposite directions with
respect to the wild-type interval, the result is scored as `unchanged'
and the boundary metabolite is not considered to be a biomarker.

@@

Here \(\bm{v}\) is the column vector of fluxes representing all
reactions in the model, \(v_i\) is the flux of the exchange reaction of
a given boundary metabolite \(i\), \(\alpha_k\) is the, possibly
negative, lower bound for reaction \(k\) and similarly \(\beta_k\) is
the, possibly negative, upper bound for reaction \(k\). The reaction
bounds are the \(V_{max}\)'s of the reactions and are part of the
metabolic network definition, typically they are set to either \(0\) or
\(\pm 1000\) if the true, biological, \(V_{max}\)'s are not known. These
values allow specification of irreversible reactions by setting the
lower bound (or upper bound) to zero. The index \(i\) specifies a single
reaction that is to be maximized, whereas the index \(k\) is used to
index the bound constraints that exist for each reaction, including
\(v_i\). \(v_i\) may be maximized in the forward (positive flux) or the
reverse (negative flux) direction if allowed by the bounds on \(v_i\).
This is indicated by the \(\pm\) symbol. \(\bm{S}\) is the stoichiometry
matrix for the metabolic network of size \(m \times r\) where \(m\) is
the number of metabolites and \(r\) the number of reactions. The
numeric, typically integer, elements of \(\bm{S}\) represent the
stoichiometry coefficients of each metabolite \(m\) in each reaction
\(r\). Consequently, for large networks, this matrix is sparse (has many
zero entries). \(Z = \bm{c}^T \bm{v}\) is the objective function defined
for the map, entailing a linear combination, defined by column vector
\(\bm{c}\), of one or more reactions. The vector \(\bm{c}\) is an
indicator of objective reactions, i.e.~it contains a value of \(1\) in
rows corresponding to fluxes that are to be included in the objective,
and zeros in all other rows. Due to the vector multiplication
\(\bm{c}^T \bm{v}\), the objective \(Z\) sums the fluxes of the
reactions that correspond to rows containing a value of \(1\) in
\(\bm{c}\). \(Z_{\text{max}}\) denotes the maximal value of the
objective function. We define \(0 \leq \phi \leq 1\) to be an arbitrary
minimal fraction of the maximal objective function value that has to be
achieved. When \(\phi=1\) we ask for the range of flux allowed for
reaction \(v_i\) while maintaining the maximal value of the linear
combination of objective fluxes. When setting \(\phi = 0\), one is
asking for the allowable flux range through reaction \(v_i\) regardless
of the value of the objective function (although the sign has to be
maintained). The latter, i.e. \(\phi=0\), is the case considered in the
method by Shlomi et al. The choice of \(\phi\) matters because when
\(\phi \neq 0\) we are requiring flux to flow through the set of
objective reactions (unless \(Z_{\text{max}} = 0\)). This may entail a
forced directionality through reactions that are required to ultimately
produce flux in the objective reactions. These additional limitations in
the freedom of the flux pattern may subsequently impact the biomarker
predictions. The approach discussed here entails predictions purely
based on the network topology and therefore sets \(\phi = 0\).

@@

We here present an implementation of the biomarker prediction method
originally proposed by Shlomi et al., programmed in Python, and
reproduce the results presented in Figure 1A-B, 2 and 3 of the original
publication. These figures concern the method's application to a simple
metabolic map and then to a set of \(17\) amino acid metabolism
disorders. Our Python module provides the biomarker prediction algorithm
and is agnostic of the map used. It can therefore be applied to other
maps than the human metabolic reconstruction considered by Shlomi et al.
Jupyter notebooks are provided that generate the illustrative map and
reproduce all results referred to in Figures 1A-B, 2 and 3 of Shlomi et
al. Our implementation is based on information given in the original
paper. However, some details critical to the reproduction were left
unexplained there and are highlighted here. We also include a more
detailed discussion of one of the amino acid IEMs to examine the
robustness of the predictions made using this approach. We feel this
adds to the utility of this reference implementation and the
understanding of the approach introduced by Shlomi et al.

@@

In this section we start with an analysis of the sensitivity of the
method which was not clearly discussed in the original publication but
is easily facilitated with our reference implementation. Subsequently,
we discuss the reproduction of Figures 1, 2 and 3 from the original
publication.

@@

PKU patients are known for elevated levels of phenylalanine and
decreased levels of tyrosine in their serum. We illustrate this briefly
in an example also considered in Figure 2 of Shlomi et al. In Figure
~\ref{fig:PKU_network}, we sketched some essential features of the
metabolic network surrounding the phenylalanine hydroxylase (PAH) enzyme
that malfunctions in PKU. Unless stated otherwise, the simulations are
performed with a medium that allows efflux through all exchange
reactions with an upper bound of \(1000\) and influx through all
exchange reactions with a lower bound of \(-1\) and we use the default
settings, including a 10\% significance threshold. The only considered
biomarkers are the \(20\) proteinogenic amino acids. We will consider
six cases in total. The results are summarized in Table
~\ref{tbl:PKU_results}.

@@

The network in Figure ~\ref{fig:Shlomi2009_figure_1}A consists of seven
metabolites, \(6\) of which (i.e.~all except M3) have exchange
reactions. We consider a hypothetical deletion of the enzyme catalyzing
the conversion of M1 into M2 coupled to the conversion of M3 into M4.
Shlomi et al. graphically provide the biomarker prediction for this
network in their Figure 1A. Here, we do so graphically and numerically
in Figure ~\ref{fig:Shlomi2009_figure_1}B and this paper is accompanied
by a Jupyter notebook reproducing the biomarker prediction. In our
analysis, all exchange reactions were given inward bounds of \(-10\) and
outward bounds of \(+1000\) so that the internal reaction bounds (of
\(\pm 1000\)) are never reached. Figure ~\ref{fig:Shlomi2009_figure_1}B
shows that all exchangeable metabolites except M5 and M7 are predicted
to be robust biomarkers.

@@

It is worthwhile to note, with an eye on the sensitivity of this
approach to the various settings, that we needed to set the medium
composition very specifically. Shlomi et al. did not report on their
chosen medium in their publication but we were able to reproduce their
results by setting all inward flux bounds to \(-1\) and all outward flux
bounds to \(1000\).

@@

\begin{figure}
\centering
\includegraphics{Figures/Reproduction_Shlomi_fig_3}
\caption{Reproduction of Figure 3 in \autocite{Shlomi2009}. (A)
Illustration of the subnetwork of Recon 1 relevant to the effect of
homocystinuria on the metabolism and transport of methionine. Nodes
represent metabolites and edges represent reactions. For simplicity,
only abbreviations of metabolite names are given in the figure. The full
names are: L-Methionine (met), S-Adenosyl-L-methionine (amet),
S-Adenosyl-L-homocysteine (ahcys), L-Homocysteine (hcys), L-Serine
(ser), L-Cystathionine (cyst), L-Cysteine (cys),
S-Adenosylmethioninamine (ametam), 5-Methylthioadenosine (5mta),
5-Methylthio-5-deoxy-D-ribose 1-phosphate (5mdr1p),
5-Methylthio-5-deoxy-D-ribulose 1-phosphate (dmdru1p),
2,3-diketo-5-methylthio-1-phosphopentane (dkmpp),
2-keto-4-methylthiobutyrate (2kmb). Metabolites with a green background
color participate in additional reactions not shown here. Reactions are
indicated by their identifiers in Recon 1 and with a corresponding
enzyme commision number if available. We highlighted the mutations for
Homocystinuria (dysfunctional CBS) and hypermethioninemia (dysfunctional
AHCY). (B) Prediction of concentration changes of methionine in
homocystinuria and S-adenosylhomocysteine hydrolase deficiency based on
the biomarker prediction algorithm.}\label{fig:Shlomi2009_figure_3}
\end{figure}
